Human Factors Influencing Visual Statistical Inference

### 2015-02-08 sociology_paper.tex
read-through up to page 8

\hh{When presenting a lineup to an observer we should also phrase the accompanying question in as general a manner as possible, effectively}
%The question that is asked of the observer should be as general as possible, 
effectively asking the observer to pick the plot that is \hh{most} different, and allowing \hh{observers} to provide their reasons for their choice. 
\hh{If lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments  in  \cite{majumder:2013}, these questions have to be phrased much more specifically, in order to make the situations more comparable. } 
%In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. 
In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked. 
In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. 
\hh{Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. 
In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one. }
% results from the turk study:
% panel:   1 10 11 12 13 14 15 16 18 19  2 20  3   4  5  6  7  8  9 
% #picks:  2  1    0  1 12    3  3    1  1   1  2   1  0 10  2  2  6 23  1

@@

 \hh{If we consider this difference to be the main objective of the visualization, a more focused question would}
 have been ``Which plot shows the biggest height difference between the two towers?'' %except that this tailors the inference to a specific feature which does not match the null hypothesis of interest.

\hh{XXX why do we need the last paragraph? It distracts from the main flow somewhat}

@@

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations \citet[see also ][]{buja:2009}.

@@

%\subsection{Demographics} 
\item{\bf Demographics:}  The experiments collected data on age, gender, education level and geographic location. Each observer self-reported gender, age in roughly five year intervals, education level as high school or less, some college courses, an undergraduate degree, some graduate courses or a graduate degree. The IP address of the computer afforded the geographical location of the subject. The purpose of collecting this information with each experiment is to examine the effect that they have on the results of visual inference -- ideally very little.

@@

%\subsection{Learning Trend} 
\item{\bf Learning Trend:} One might expect that as an observer evaluates more lineups they may become more skillful in their evaluations. Each participant in  experiments~5, 6 and~7 was asked to evaluate a block of  ten lineups of the same type of data plot. The ten lineups were randomly chosen from the lineups  produced for each study. Before evaluating the first lineup, the observer needs to read instructions and become accustomed to the type of plot used in the lineup. In subsequent lineups  the type of plot is the same. It is possible that the observer becomes more skillful at recognizing the most different plot, either by more often detecting the actual data plot or more quickly reporting their choice. These two ways of measuring learning trend are evaluated on data collected from experiments~5, 6 and~7.

@@

%\subsection{Location of Actual Data Plot in the Lineup}
\item{\bf Location of Actual Data Plot in the Lineup} For all of the lineups used in the experiments a $5\times 4$ grid of 20 points is used. A random number generator is used to determine the position where the actual data plot is placed in each lineup. Eye tracking experiments \citep{zhao:2012} suggest that some observers traverse lineups in horizontal direction, while others have a vertical up-and-down approach to their search. Almost universal, observers start at the top left of a lineup. This leads to some concern that observers might be able to more easily identify the actual data plot, if it is placed at the top left of a lineup than when it the actual data is placed at the bottom right.
%With the help of eye-tracking equipment \citet{zhao:2012} tracked the observers' eyes to see their approach in searching the plots in a lineup. The results suggest that many people read lineups from left to right, some read from upward to downward. 
%There is some concern that if the actual data plot is located in the top left of the lineup, observers may be able to identify it more easily than when the actual data plot is at the bottom right. 
Ideally, this does not happen. Experiment~9  is designed to allow us to investigate the effects of location on detection. Five different locations in the lineup were used to assess how fast and accurately observers identified the actual data plot.

@@

%\subsection{Individual Skill or Ability} 
\item{\bf Individual Skill or Ability:} Each person may have different visual skill sets, and aptitude for reading statistical plots. This can be examined by having the same lineups evaluated by multiple subjects, so that it can be seen if some subjects consistently detect the actual data plot more often than others.  
 %For example, in a controlled experimental study \citet{zhao:2012}  noticed that some people spent a lot of time to decide no matter whether the lineup is difficult or easy while some simply glanced at lineups to make a decision. This influences the response of the observer. Also subject specific variation in the power of visual test is observed in \cite{majumder:2013}. 
\end{itemize}

@@

All the ten experiments shown in Table \ref{tbl:visual_stat}, had demographic information collected from subjects, as follows:

\begin{enumerate} \leftmargin 5cm  \itemsep 0in
\item Age group, with categories set to be 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60.
\item Gender, male or female.
\item Education level, with levels being high school or less, some under grad course, under graduate degree, some graduate courses, and graduate degree.
\item Geographical location, collected from the IP address of the participants' computer, as latitude, longitude, city and country. 
\end{enumerate}

@@

Let $Y_{ij}$ denote the response from observer $i$ on a lineup $j$, with $Y_{ij} =1$ if the actual data plot is chosen, otherwise $Y_{ij} =0$. The factors are examined in association with the observer's response using a logistic regression with random effects terms:
\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}
where $\pi_{ij}=  E(Y_{ij})$ is the probability that  observer $i$ picks the actual data plot from lineup $j$, %We model this in a generalized mixed effects model of the form
%\begin{equation} \label{eqn:demographic_response}
%g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
%\end{equation}
 $\mu$ is an overall average probability for picking out the data plot from a lineup, $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender $l(i)$, education level $m(i)$ and country name $s(i)$ respectively for observer $i$. The term $\ell_j$ is a random intercept predicting lineup difficulty level and we assume  $\ell_j \sim N(0, \sigma_\ell^2)$, and $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.

In addition, we model the  time  an observer takes to identify a panel from a lineup. Let $Z_{ij}$ denote the logarithm of time taken for  observer $i$ to evaluate  lineup $j$. Let $\mu_{ij}=  E(Z_{ij})$ be the average of the (log) time taken by  observer $i$ to pick a panel from lineup $j$. We model this in a mixed effects model of the same structure as model~(\ref{eqn:demographic_response}) given as:
%\hh{
\begin{equation} \label{eqn:demographic_time}
Z_{ij} = \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j+ \epsilon_{ij},  
\end{equation}
where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup, $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are as described in model \eqref{eqn:demographic_response}, $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup, with $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ij} \sim N(0, \sigma^2)$.

>>>>

\hh{When presenting a lineup to an observer we should also phrase the accompanying question in as general a manner as possible, }
%The question that is asked of the observer should be as general as possible, 
effectively asking the observer to pick the plot that is \hh{most} different, and allowing \hh{observers} to provide their reasons for their choice. 
\hh{However, if lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments  in  \cite{majumder:2013}, these questions have to be phrased much more specifically, in order to make the situations more comparable. } 
%In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. 
In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked. 
In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. 
\hh{Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. 
In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one. }
% results from the turk study:
% panel:   1 10 11 12 13 14 15 16 18 19  2 20  3   4  5  6  7  8  9 
% #picks:  2  1    0  1 12    3  3    1  1   1  2   1  0 10  2  2  6 23  1

@@

 \hh{If we consider this difference to be the main objective of the visualization, a more focused question, that aligns more with a hypothesis test on the difference  in electoral votes  between parties, would}
 have been ``Which plot shows the biggest height difference between the two towers?'' %except that this tailors the inference to a specific feature which does not match the null hypothesis of interest.

@@

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations \citep[see also ][]{buja:2009}.

@@

%\subsection{Demographics} 
\item{\bf Demographics:}  \hh{During the experiments } data on age, gender, education level and geographic location \hh{was collected}. Each observer self-reported gender, age in roughly five year intervals, education level as high school or less, some college courses, an undergraduate degree, some graduate courses or a graduate degree. The IP address of the computer afforded the geographical location of the subject. The purpose of collecting this information with each experiment is to examine the effect that they have on the results of visual inference -- ideally very little.

@@

%\subsection{Learning Trend} 
\item{\bf Learning Trend:} One might expect that as an observer evaluates more lineups they  become more skillful in their evaluations. Each participant in  experiments~5, 6 and~7 was asked to evaluate a block of  ten lineups of the same type of data plot. The ten lineups were randomly chosen from the lineups  produced for each study. Before evaluating the first lineup, the observer needs to read instructions and become accustomed to the type of plot used in the lineup. In subsequent lineups  the type of plot is the same. It is possible that the observer becomes more skillful at recognizing the most different plot, either by more often detecting the actual data plot or more quickly reporting their choice. These two ways of measuring learning trend are evaluated on data collected from experiments~5, 6 and~7.

@@

%\subsection{Location of Actual Data Plot in the Lineup}
\item{\bf Location of Actual Data Plot in the Lineup} For all of the lineups used in the experiments a $5\times 4$ grid of 20 \hh{panels} is used. A random number generator is used to determine the position where the actual data plot is placed in each lineup. Eye tracking experiments \citep{zhao:2012} suggest that some observers traverse lineups in horizontal direction, while others have a vertical up-and-down approach to their search. Almost universally, observers start at the top left of a lineup. This leads to some concern that observers might be able to more easily identify the actual data plot, if it is placed at the top left of a lineup than when it the actual data is placed at the bottom right.
%With the help of eye-tracking equipment \citet{zhao:2012} tracked the observers' eyes to see their approach in searching the plots in a lineup. The results suggest that many people read lineups from left to right, some read from upward to downward. 
%There is some concern that if the actual data plot is located in the top left of the lineup, observers may be able to identify it more easily than when the actual data plot is at the bottom right. 
Ideally, this does not happen. Experiment~9  is designed to allow us to investigate the effects of location on detection. Five different locations in the lineup were used to assess how fast and accurately observers identified the actual data plot.

@@

%\subsection{Individual Skill or Ability} 
\item{\bf Individual Skill or Ability:} Each person may have different  aptitude for reading statistical plots, and their visual skill sets might be differently developed. We \hh{We can evaluate the effect of individuals' skills and abilities}  by letting multiple subjects evaluate the same lineups, which allows us to estimate, if some subjects consistently detect the actual data plot more often than others.  
 %For example, in a controlled experimental study \citet{zhao:2012}  noticed that some people spent a lot of time to decide no matter whether the lineup is difficult or easy while some simply glanced at lineups to make a decision. This influences the response of the observer. Also subject specific variation in the power of visual test is observed in \cite{majumder:2013}. 
\end{itemize}

@@

For all  ten experiments shown in Table \ref{tbl:visual_stat}, the following demographic information was collected from subjects:

\begin{enumerate} \leftmargin 5cm  \itemsep 0in
\item {\it Age group}, with categories set to be 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60.
\item {\it Gender}, male or female.
\item {\it Education level}, with levels being high school or less, some undergraduate courses, undergraduate degree, some graduate courses, and graduate degree.
\item {\it Geographical location}, collected from the IP address of the participants' computer, as latitude, longitude, city and country. 
\end{enumerate}

@@

Let $Y_{ij}$ denote the response from observer $i$ on a lineup $j$, with $Y_{ij} =1$ if the actual data plot is chosen, otherwise $Y_{ij} =0$. The factors are examined in association with the observer's response using a logistic regression with random effect terms:
\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}
where $\pi_{ij}=  E(Y_{ij})$ is the probability that  observer $i$ picks the actual data plot from lineup $j$, %We model this in a generalized mixed effects model of the form
%\begin{equation} \label{eqn:demographic_response}
%g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
%\end{equation}
 $\mu$ is an overall population average, %probability for picking out the data plot from a lineup HH: mu is not a probability, it is on the scale of the link function
 $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender $l(i)$, education level $m(i)$ and country name $s(i)$, respectively for observer $i$. The term $\ell_j$ is a random intercept predicting lineup difficulty level and we assume independence and normality of the errors, i.e.\ $\ell_j \sim N(0, \sigma_\ell^2)$.  $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.

Similarly, we model the  time  an observer takes to identify a panel from a lineup. Let $Z_{ij}$ denote the logarithm of time taken for  observer $i$ to evaluate  lineup $j$. Let $\mu_{ij}=  E(Z_{ij})$ be the average of the (log) time taken by  observer $i$ to pick a panel from lineup $j$. We model this in a mixed effects model of the same structure as model~(\ref{eqn:demographic_response}) given as:
%\hh{
\begin{equation} \label{eqn:demographic_time}
Z_{ij} = \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j+ \epsilon_{ij},  
\end{equation}
where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup, $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are as described in model \eqref{eqn:demographic_response}, $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup, with $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ij} \sim N(0, \sigma^2)$.
